A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
move the table up, to force latex to put it into a sensible place
more citations

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronisation is not easy\cfbolz{citation needed
:-). would be cool if you could find something}.  They are
non-composable, have overhead, may deadlock, limit scalability, and
overall add a lot of complexity\cfbolz{same here, really}. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic blocks}\cfbolz{and here}.

@@

\begin{table*}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & -    & -  & -    & -  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

>>>>

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  & o    & -  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronisation is not
easy~\cite{christopher10,victor11,shan08}. They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic
  blocks}~\cite{tim03,tim05}.

@@

@@

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
  transactional memory vs. locks in practice. In \emph{Proceedings of
    the twenty-third annual ACM symposium on Parallelism in algorithms
    and architectures} (SPAA '11). ACM, New York, NY, USA

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. Is transactional programming actually
  easier?. \emph{SIGPLAN} Not. 45, 5 (January 2010), 47-56.

\bibitem{tim03}
  Tim Harris and Keir Fraser. 2003. Language support for lightweight
  transactions. \emph{In Proceedings of the 18th annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '03).

\bibitem{tim05}
  Tim Harris, Simon Marlow, Simon Peyton-Jones, and Maurice
  Herlihy. 2005. Composable memory transactions. \emph{In Proceedings
    of the tenth ACM SIGPLAN symposium on Principles and practice of
    parallel programming} (PPoPP '05).

\bibitem{shan08}
  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. Learning
  from mistakes: a comprehensive study on real world concurrency bug
  characteristics. \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
  329-339.
